---
title: 'Class17: Mini Project'
author: 'Lawrence Adhinatha (PID: A16035882)'
date: "2023-03-07"
output: github_document
---

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> Q1. What column details the total number of people fully vaccinated? 

vax$persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area? 

vax$zip_code_tabulation_area

> Q3. What is the earliest date in this dataset? 

```{r}
min(vax$as_of_date)
vax$as_of_date[1]
```

2021-01-05

> Q4. What is the latest date in this dataset? 

```{r}
max(vax$as_of_date)
vax$as_of_date[nrow(vax)]
```

2023-02-28


Using the skimr package to get an overview of the dataset: 
```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 

13 numeric columns

> Q6. Note that there are “missing values” in the dataset. How many NA values are there in the persons_fully_vaccinated column? 

```{r}
n.missing <- sum(is.na(vax$persons_fully_vaccinated))
n.missing
```
16525 NA values

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
round((n.missing/nrow(vax))*100, 2)
```

8.29% are missing

> Q8. [Optional]: Why might this data be missing?

The data may be redacted for legal and privacy measures.

## Working with dates

Lubridate package: makes working with dates and times in R much less of a pain. Let's have a first play with this package here. 

```{r}
library(lubridate)
```


```{r}
today()
```

We can now magically do math with dates; 

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
today() - vax$as_of_date[1]
```


How old am I

```{r}
today() - ymd("2001-07-24")

```


```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]

```

How many days does the dataset span? 

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[nrow(vax)]
```
7 days have passed since the last update of the dataset (it updates weekly). 

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```

```{r}
library(dplyr)
n_distinct(vax$as_of_date)
```

## Working with ZIP codes

```{r}
library(zipcodeR)
```

```{r}
# getting the centroid of the ZIP code
geocode_zip('92037')
```

```{r}
# calculating the distance (in miles) between two ZIP code centroids
zip_distance('92037','91387')
```

```{r}
# pulling data from selected ZIP codes
reverse_zipcode(c('92037','91387') )
```

```{r}
zipdata <- reverse_zipcode(vax$zip_code_tabulation_area)
```


## Focusing on the San Diego area

```{r}
# Subset to San Diego county only areas
sd <- vax[vax$county=="San Diego",]
nrow(sd)
```

With dplyr: 

```{r}
library(dplyr)
sd <- filter(vax, county == "San Diego")

nrow(sd)
```

dplyr proves convenient when subsetting across multiple criteria (i.e. all sd county areas with pop>10000)

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                  age5_plus_population > 10000)
nrow(sd.10)
```

> How many zip codes are we dealing with?

```{r}
n_distinct(sd.10$zip_code_tabulation_area)
```

> Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
n_distinct(sd$zip_code_tabulation_area)
```

107 distinct zip codes


> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 

```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]
```

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-11-15”? 

```{r}
vax$as_of_date[nrow(vax)]
```

```{r}
sd.current <- filter(vax, county == "San Diego" &
                       as_of_date == "2023-02-28")

mean(sd.current$percent_of_population_fully_vaccinated, na.rm=TRUE) * 100

```
74%


> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?

```{r}
hist(sd.current$percent_of_population_fully_vaccinated, main = "Histogram of Vaccination Rates Across San Diego County", xlab = "Percent of People Fully Vaccinated as of 2023-02-28")
```

```{r}
library(ggplot2)

ggplot(sd.current) +
  aes(x=percent_of_population_fully_vaccinated) + 
  geom_histogram() +
  labs(title = "Histogram of Vaccination Rates Across San Diego County", subtitle = "As of 2023-02-28", x = "Percent of Population Fully Vaccinated in a ZIP code area", y = "Count (ZIP code areas)")

  
```

## Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```


> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ucplot <- ggplot(ucsd) + 
  aes(as_of_date, percent_of_population_fully_vaccinated) + 
  geom_point() + 
  geom_line(group=1) + 
  ylim(c(0,1)) + 
  labs(title = "Vaccination Rate for La Jolla CA 92037", x = "Date", y = "Percent Vaccinated")
ucplot
```

## Comparing to similar sized areas

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on the most recent date. 

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 &
                   as_of_date == "2023-02-28")
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
avg <- mean(vax.36$percent_of_population_fully_vaccinated)
avg
```
72.13%

```{r}
ucplot_line <- ucplot + geom_hline(yintercept=avg, col="red", linetype=2)
ucplot_line
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-02-28”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```


> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() + 
  labs(x = "Percent Vaccinated") + 
  xlim(c(0,1))
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
x <- filter(vax.36, zip_code_tabulation_area %in% c("92109", "92040"))
x$percent_of_population_fully_vaccinated 
avg

x$percent_of_population_fully_vaccinated < avg
```
Both ZIP code areas have values below the average value for all ZIP codes.

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only areas with a population over 36k are shown.") +
  geom_hline(yintercept = avg, linetype=2)
```



















